Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums

### 2019-06-08 papers/halfdisk/halfdisk.tex
Fixed a few typos
readthrough

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
The corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:construction}.

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  In particular, by exploiting the connection with one-dimensional orthogonal polynomials we are able to construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D, and it is clear from the construction in this paper that discretizations of spherical gradients and Laplacian's are sparse on half-spheres and other suitable sub-components of the sphere. The resulting sparsity in high-polynomial degree discretisations presents an attractive alternative to methods based on bijective mappings (e.g., \cite{DGShallowWater,FEMShallowWater}).
%
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in turbulent pipe flow.

@@

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition. First we introduce notation for two families of univariate OPs
\begin{definition}\label{def:OPconstruction}
Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $\jac_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y \jac_n^{(b,b)}(y) &= \delta_n^{(b)} \jac_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} \jac_n^{(b,b)}(y) + \delta_{n-1}^{(b)} \jac_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align*}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align*}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a function's expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \mapsto x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.

$J_x^{(a,b)}, J_y^{(a,b)}$ are banded-block-banded matrices:

\begin{definition}
A block matrix $A$ with blocks $A_{i,j}$ has block-bandwidths $(L,U)$ if $A_{i,j} = 0$ for $- L \leq j-i \leq U$, and sub-block-bandwidths $(\lambda, \mu)$ if all blocks $A_{i,j}$ are banded with bandwidths $(\lambda,\mu)$. A matrix where the block-bandwidths and sub-blockbandwidths are small compared to the dimensions is referred to as a banded-block-banded matrix. 
\end{definition}

@@

We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$. Since $\{\hdop^{(a, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j}^y = {\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^{-2}_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	&\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} \\
	&\quad \quad = \normgenjac^{(a, b+1)} \: \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}} \\
	&\quad \quad = \normgenjac^{(a, b+1+\half(k+j))} \: \normgenjac^{(a, b+1)} \: \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}}.
\end{align*}
Now, using (\ref{eqn:derivativeofweightjac}), integration-by-parts, and noting that the weight $\jacw^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}} &= \int_{-1}^1 \: \jac_k^{(b,b) \: \prime}(y) \: \jac_{j}^{(b+1,b+1)}(y) \: \jacw^{(b+1)}(y) \: \D y \\
	&= - \int_{-1}^1 \: \jac_k^{(b,b)}(y) \: \ddy [\jacw^{(b+1)}(y) \: \jac_{j}^{(b+1,b+1)}(y)] \: \D y \\
	&= - \int_{-1}^1 \: \jac_k^{(b,b)} \: [ \jac_{j}^{(b+1,b+1) \: \prime} \: \jacw^{(b+1)} - 2 b \: y \: \jac_{j}^{(b+1,b+1)} \:\jacw^{(b)} ] \: \D y \\
	&= - \: \normjac^{(b)} \: \ip< \jac_k^{(b,b)}, \: \jacw^{(1)} \: \jac_{j}^{(b+1,b+1) \: \prime} -2b \:y \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b)}}
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} = \delta_{n,m+1}
\end{align*}
showing the only possible non-zero coefficient is when $m=n-1, j=k-1$. Finally, we simply have that 
\begin{align*}
	c_{n-1,k-1}^y &= \normgenjac^{(a, b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{k-1}^{(b+1,b+1)} >_{\jacw^{(b+1)}}.
\end{align*}

We next consider the case for the operator $D_x^{(a,b)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a+1, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. As before, we can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-1, k-2$ and $0 \le j \le m$. First, note that
\begin{align*}
	c_{m,j}^x = {\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{\hdop^{(a+1, b+1)}_{m,j}}^{-2}_{W^{(a+1, b+1)}}}.
\end{align*}
Second, note that
\begin{align}
	\rho'(x) = -x \: \rho(x)^{-2}. \label{eqn:rhoderivative}
\end{align}
Now, using the change of variable $t= \frac{y}{\rho(x)}$ and (\ref{eqn:rhoderivative}), we have that
\begin{align*}
	&\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \normgenjac^{(a, b+\half)} \: \normjac^{(b+1)} \: \Big\{ \ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime} (x), \: \rho(x)^{k+j+2} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} \\ 
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \cdot \: \ip< \jac_k^{(b,b)}, \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< \jac_k^{(b,b)}, \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< y \: \jac_k^{(b,b) \: \prime} (y), \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}} \Big\}
\end{align*}

>>>>

%\bibliographystyle{apalike}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
The corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  and \cite{vasil2016tensor} for equations on the disk results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  In particular, by exploiting the connection with one-dimensional orthogonal polynomials we are able to construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D, and it is clear from the construction in this paper that discretizations of spherical gradients and Laplacian's are sparse on half-spheres and other suitable sub-components of the sphere. The resulting sparsity in high-polynomial degree discretisations presents an attractive alternative to methods based on bijective mappings (e.g., \cite{DGShallowWater,FEMShallowWater,boyd2005sphere}).
%
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in turbulent pipe flow.

@@

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definition. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:OPconstruction}
Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $\jac_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y \jac_n^{(b,b)}(y) &= \delta_n^{(b)} \jac_{n+1}^{(b,b)}(y) + \delta_{n-1}^{(b)} \jac_{n-1}^{(b,b)}(y).
\end{align}
 We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align*}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align*}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a function's expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \mapsto x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis. Further, note that $J_x^{(a,b)}, J_y^{(a,b)}$ are banded-block-banded matrices:

\begin{definition}
A block matrix $A$ with blocks $A_{i,j}$ has block-bandwidths $(L,U)$ if $A_{i,j} = 0$ for $- L \leq j-i \leq U$, and sub-block-bandwidths $(\lambda, \mu)$ if all blocks $A_{i,j}$ are banded with bandwidths $(\lambda,\mu)$. A matrix where the block-bandwidths and sub-block-bandwidths are small compared to the dimensions is referred to as a banded-block-banded matrix. 
\end{definition}

@@

We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$. Since $\{\hdop^{(a, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^y \: \hdop^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j}^y = {\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^{-2}_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \normgenjac^{(a, b+1)} \: \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \\
	&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}} \\
	&= \normgenjac^{(a, b+1+\half(k+j))} \: \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \\
		&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}}.
\end{align*}
Now, using (\ref{eqn:derivativeofweightjac}), integration-by-parts, and noting that the weight $\jacw^{(b)}$ is a polynomial of degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\normjac^{(b+1)} \: \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b+1)}} &= \int_{-1}^1 \: \jac_k^{(b,b) \: \prime}(y) \: \jac_{j}^{(b+1,b+1)}(y) \: \jacw^{(b+1)}(y) \: \D y \\
	&= - \int_{-1}^1 \: \jac_k^{(b,b)}(y) \: \ddy [\jacw^{(b+1)}(y) \: \jac_{j}^{(b+1,b+1)}(y)] \: \D y \\
	&= - \int_{-1}^1 \: \jac_k^{(b,b)} \: [ \jac_{j}^{(b+1,b+1) \: \prime} \: \jacw^{(b+1)} - 2 b \: y \: \jac_{j}^{(b+1,b+1)} \:\jacw^{(b)} ] \: \D y \\
	&= - \: \normjac^{(b)} \: \ip< \jac_k^{(b,b)}, \: \jacw^{(1)} \: \jac_{j}^{(b+1,b+1) \: \prime} -2b \:y \: \jac_{j}^{(b+1,b+1)} >_{\jacw^{(b)}}
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} = \delta_{n,m+1},
\end{align*}
showing that the only possible non-zero coefficient is when $m=n-1, j=k-1$. Finally,
\begin{align*}
	c_{n-1,k-1}^y &= \ip< \jac_k^{(b,b) \: \prime}, \: \jac_{k-1}^{(b+1,b+1)} >_{\jacw^{(b+1)}}.
\end{align*}

We next consider the case for the operator $D_x^{(a,b)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^x \: \hdop^{(a+1, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. As before, we can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2$, $j = k, k-1, k-2$ and $0 \le j \le m$. First, note that
\begin{align*}
	c_{m,j}^x = {\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{\hdop^{(a+1, b+1)}_{m,j}}^{-2}_{W^{(a+1, b+1)}}}.
\end{align*}
Second, note that
\begin{align}
	\rho'(x) = -x \: \rho(x)^{-1}. \label{eqn:rhoderivative}
\end{align}
Now, using the change of variable $t= \frac{y}{\rho(x)}$ and (\ref{eqn:rhoderivative}), we have that
\begin{align*}
	&\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 
	&= \normgenjac^{(a, b+\half)} \: \normjac^{(b+1)} \: \Big\{ \ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime} (x), \: \rho(x)^{k+j+2} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} \\ 
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \cdot \: \ip< \jac_k^{(b,b)}, \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}}  \nonumber \\ 
	& \quad - k \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< \jac_k^{(b,b)}, \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}}  \nonumber \\ 
	& \quad + \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} 
		\: \ip< y \: \jac_k^{(b,b) \: \prime} (y), \: \jac_j^{(b+1,b+1)} >_{\jacw^{(b+1)}} \Big\}
\end{align*}

We can see  from each term's second factor that the above is zero for $j < k-2$. Now, using (\ref{eqn:derivativeofweightgenjac}), integration-by-parts, and noting that the weight $\genjacw^{(a,b)}$ is a polynomial degree $a+2b$ and vanishes at the limits of the integral for positive parameters $a,b$, we have that
\begin{align*}
	&\normgenjac^{(a, b+\half)} \: \ip< x \: \genjacnmk^{(a, b+k+\half) \: \prime}, \: \rho(x)^{k+j+2} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}}  \\
	&= \int_0^1 \: \genjacnmk^{(a, b+k+\half) \: \prime} \: \genjacmmj^{(a+1, b+1+j+\half)} \: \genjacw^{(a+1, b+1+\half(k+j+1))} \: \D x \\
	&= - \int_0^1 \: \genjacnmk^{(a, b+k+\half)} \: \ddx \Big[ \genjacmmj^{(a+1, b+1+j+\half)} \: \genjacw^{(a+1, b+1+\half(k+j+1))} \Big] \: \D x \\
	&= - \int_0^1 \: \genjacnmk^{(a, b+k+\half)} \: \Big\{ \genjacw^{(a+1, b+1+\half(k+j+1))} \: \genjacmmj^{(a+1, b+1+j+\half) \: \prime} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (a+1) \: \genjacw^{(a, b+1+\half(k+j+1))} \: \genjacmmj^{(a, b+1+j+\half)} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad -2(b+1+\half(k+j+1)) \: \genjacw^{(a+2, b+\half(k+j+1))} \: \genjacmmj^{(a, b+1+j+\half)} \Big\} \: \D x \\
	&= - \: \normgenjac^{(a,b+k+\half)} \:  \Big\{ \ip<\genjacnmk^{(a, b+k+\half)}, \: \genjacw^{(1, 1+\half(j-k))} \: \genjacmmj^{(a+1, b+1+j+\half) \: \prime}>_{\genjacw^{(a, b+k+\half)}} \\
		&\quad \quad \quad + \ip<\genjacnmk^{(a, b+k+\half)}, \: (a+1) \: \genjacw^{(0,1+\half(j-k))} \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}} \\
		&\quad \quad \quad - \ip<\genjacnmk^{(a, b+k+\half)}, \: 2(b+1+\half(k+j+1)) \: \genjacw^{(2, \half(j-k))} \: \genjacmmj^{(a, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}} \Big\}
\end{align*}
which is zero for $m < n - 2$, and further that
\begin{align*}
	&\normgenjac^{(a, b+\half)} \: \ip< x^2 \: \genjacnmk^{(a, b+k+\half)}(x), \: \rho(x)^{k+j} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+\half)}} \\
	&= \normgenjac^{(a, b+k+\half)} \: \ip< \genjacnmk^{(a, b+k+\half)}(x), \: \genjacw^{(2, \half(j-k))} \: \genjacmmj^{(a+1, b+1+j+\half)} >_{\genjacw^{(a, b+k+\half)}}
\end{align*}
which is also zero for $m < n - 2$.
